Title: Class size paradox
Author: Dr. Glasbrenner
Author_Image: https://www.gravatar.com/avatar/49802fdfa5a0e63b3d932a5179d41c1e
Date: 2018-04-10 15:00
Tags: reading-nograyout
Slug: class-size-paradox
Summary: A practical example that shows how to use PMFs to resolve a paradox.
Show_summary: true
Show_link: true

```{r setup, include = FALSE}
icon_pdf <- '<i class="fas fa-file-pdf" data-fa-transform="grow-8"></i>&nbsp;'
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.width = 6, fig.alt = "",
                      fig.asp = 0.618, dpi = 120, dev = "svg",
                      out.width = "60%", fig.align = "center", cache = TRUE)
library(tidyverse)
```

::: {.no-bullets}

*   [`r icon_pdf` Download this reading][reading-pdf]

:::

```r
library(tidyverse)
```

**The following example assumes you have already read the standard reading on [_Probability mass functions_][pmf-reading].**

Let's consider another PMF computation that illustrates something that we may call the "class size paradox".

At many American colleges and universities, the student-to-faculty ratio is about 10:1.
But students are often surprised to discover that their average class size is bigger than 10.
There are two reasons for the discrepancy:

*   Students typically take 4 – 5 classes per semester, but professors often teach 1 or 2.
*   The number of students who enjoy a small class is small, but the number of students in a large class is (unsurprisingly) large.

The first effect is obvious, at least once it is pointed out; the second is more subtle.
Let's look at an example.
Suppose that a college offers 65 classes in a given semester, with the following distribution of sizes:

| size      | count  |
| --------- | ------ |
| 5--9      | 8      |
| 10--14    | 8      |
| 15--19    | 14     |
| 20--24    | 4      |
| 25--29    | 6      |
| 30--34    | 12     |
| 35--39    | 8      |
| 40--44    | 3      |
| 45--49    | 2      |

If you ask the Dean for the average class size, he or she would construct a PMF, compute the mean, and report that the average class size is 23.7.
Here's the code:

```{r class-sizes-setup-and-average, eval = FALSE}
class_sizes <- tribble(
  ~`class size`, ~count,
  7, 8,
  12, 8,
  17, 14,
  22, 4,
  27, 6,
  32, 12,
  37, 8,
  42, 3,
  47, 2)

class_sizes %>%
  summarize(`Average class size` = round(weighted.mean(`class size`, count), 1))
```

```{r class-sizes-setup-and-average-table, echo = FALSE}
class_sizes <- tribble(
  ~`class size`, ~count,
  7, 8,
  12, 8,
  17, 14,
  22, 4,
  27, 6,
  32, 12,
  37, 8,
  42, 3,
  47, 2)

class_sizes %>%
  summarize(
    `Average class size` = round(weighted.mean(`class size`, count), 1)) %>%
  knitr::kable(format = "html")
```

But if you survey a group of students, ask them how many students are in their classes, and compute the mean, you would think the average class was bigger.
Let's see how much bigger.

First, let's compute the distribution as observed by students, where the probability associated with each class size is "biased" by the number of students in the class.
For each class size we multiply the probability by the number of students who observe that class size.
The result is a new PMF that represents the biased distribution:

```{r compute-observed-class-size}
class_sizes2 <- class_sizes %>%
  mutate(actual = count / sum(count)) %>%
  mutate(observed = (actual * `class size`) / sum(actual * `class size`)) %>%
  gather(key = distribution, value = PMF, actual:observed)
```

Now we can plot the actual and observed distributions together:

```{r actual-vs-observed-class-size, eval = FALSE}
class_sizes2 %>%
  ggplot(
    mapping = aes(x = `class size`, y = PMF, fill = distribution,
                  color = distribution)) +
  geom_col(position = "identity", alpha = 0.5)
```

```{r actual-vs-observed-class-size-labeled, echo = FALSE}
class_sizes2 %>%
  ggplot(
    mapping = aes(x = `class size`, y = PMF, fill = distribution,
                  color = distribution)) +
  geom_col(position = "identity", alpha = 0.5) +
  labs(
    title = "Distributions of actual and observed class sizes",
    x = "Class size", y = "PMF")
```

As we see in the above figure, the biased distribution corresponds to fewer small classes and more large ones.
The mean of the biased distribution is,

```{r observed-class-mean, eval = FALSE}
class_sizes2 %>%
  group_by(distribution) %>%
  summarize(`Average class size` = round(weighted.mean(`class size`, `PMF`), 1))
```

```{r observed-class-mean-table, echo = FALSE}
class_sizes2 %>%
  group_by(distribution) %>%
  summarize(
    `Average class size` = round(weighted.mean(`class size`, `PMF`), 1)) %>%
  knitr::kable(format = "html")
```

which is almost 25% higher than the actual mean.

It is also possible to invert this operation.
Suppose you want to find the distribution of class sizes at a college, but you can't get reliable data from the Dean.
An alternative is to choose a random sample of students and ask how many students are in their classes.
The result would be biased for the reasons we've just seen, but you can use it to estimate the actual distribution.
Here's the code that can be used to unbias a PMF:

```{r unbias-pmf, eval = FALSE}
class_sizes2 %>%
  spread(key = distribution, value = PMF) %>%
  mutate(
    unbiased = (observed * 1 / `class size`) /
      sum(observed * 1 / `class size`)) %>%
  select(`class size`, count, observed, unbiased, actual)
```

```{r unbias-pmf-table, echo = FALSE}
class_sizes2 %>%
  spread(key = distribution, value = PMF) %>%
  mutate(
    unbiased = (observed * 1 / `class size`) /
      sum(observed * 1 / `class size`)) %>%
  select(`class size`, count, observed, unbiased, actual) %>%
  knitr::kable(format = "html")
```

It's similar to before; the only difference is that it divides each probability by the class size instead of multiplying.

## Credits

::: {.license}
This work, *Class size paradox*, is a derivative of [Allen B. Downey, "Chapter 3 Probability mass functions" in *Think Stats: Exploratory Data Analysis*, 2nd ed. (O'Reilly Media, Sebastopol, CA, 2014)][think-stats-2], used under [CC BY-NC-SA 4.0][cc-4-by-nc-sa].
*Class Size Paradox* is licensed under [CC BY-NC-SA 4.0][cc-4-by-nc-sa] by James Glasbrenner.
:::

[pmf-reading]:   /materials/representing-distributions-pmf/
[reading-pdf]:   /doc/class_size_paradox.pdf
[think-stats-2]: http://a.co/grOJGrv 
[cc-4-by-nc-sa]: http://creativecommons.org/licenses/by-nc-sa/4.0/
